\vspace{1.0cm} \LARGE \bf Integrating influenza antigenic dynamics with molecular evolution

### 2012-04-03 flux.tex
Reverted.
Moving BibTeX references into the repo.
Cleaning up Gytis's references.
Including some initial results on MDS error and antigenic drift.

% cite package, to clean up citations in the main text. Do not remove.
%\usepackage{cite}
%\renewcommand\citeleft{(}
%\renewcommand\citeright{)}
%\renewcommand\citeform[1]{\textsl{#1}}

@@

\author[1]{Trevor Bedford}
\author[2,3,4]{Marc A. Suchard}
\author[5]{Philippe Lemey}
\author[1]{Gytis Dudas}
\author[6]{Colin Russell}
\author[6,7]{Derek Smith}
\author[1,8]{Andrew Rambaut}

@@

Mutations to the HA1 region of the hemagglutinin (HA) protein are thought to drive the majority of antigenic drift in the influenza virus \cite{Nelson07NatRevGenet}. 
Experimental characterization of antigenic phenotype is possible through the hemagglutination inhibition (HI) assay \cite{Hirst43}, which measures the cross-reactivity of one virus strain to serum raised against another strain. 
Here is a new sentence.
Sera from older strains react poorly with more recent viruses resulting in new strains having a transmission advantage over established strains. 
The results of many HI assays across a multitude of virus strains can be combined to yield a two-dimensional map, representing antigenic similarity and distance as an easily visualized and quantified measure \cite{Smith04}. 
The antigenic map of influenza A (H3N2) has shown largely linear movement of the influenza virus population since its introduction in 1968. However, evolution of antigenic phenotype appears punctuated with periods of stasis interspersed by periods of more rapid innovation, while genetic evolution appears more continuous \cite{Smith04}, suggesting that a relatively small number of genetic changes or combinations of genetic changes may drive changes in antigenic phenotype. 
The process of antigenic drift results in the rapid turnover of the virus population. 
Although mutation occurs rapidly, standing genetic diversity is low and phylogenetic analysis shows a characteristically `spindly' tree with a single predominant trunk lineage and transitory side branches that persist for only 1--5 years \cite{Fitch97}.

@@

%%% Figure 1: mds %%%
\begin{figure}[tb]
	\centering		
	\includegraphics[width=\textwidth]{figures/mds}
	\caption{\textbf{Antigenic locations of influenza H3N2 and H1N1.} (A) and (B) Antigenic maps showing the mean posterior location of 338 strains of H3N2 influenza and 243 strains of H1N1 influenza.  
	The map has been oriented so that the primary axis of variation lies along the $x$-axis (AG1), with the $y$-axis (AG2) orthogonal to this axis.  
	(C) and (D) Antigenic location along the primary axis of variation (AG1) vs.\ year of virus isolation.  
	The dashed lines show the relationship of between time and AG1 with a slope of \tbc{XXX} for H3N2 and \tbc{XXX} for H1N1.  
	(E) and (F) Antigenic location along the secondary axis of variation (AG2) vs.\ year of virus isolation.  
	Antigenic units represent two-fold dilutions of the HI assay, and strains have been colored based on year of isolation.  
	\tbc{Decide whether to just show mean posterior or show distributions.}
	} 
	\label{mds} 
\end{figure}

@@

Antigenic data for influenza A (H1N1) was collected from sources... \tbc{ask Gytis}.

@@

%%% REFERENCES %%%
\bibliographystyle{plos}
\bibliography{/Users/bedfordt/Documents/bedford}

>>>>

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}
%\renewcommand\citeleft{(}
%\renewcommand\citeright{)}
%\renewcommand\citeform[1]{\textsl{#1}}

@@

\author[1]{Trevor Bedford}
\author[2,3,4]{Marc A. Suchard}
\author[1]{Gytis Dudas}
\author[5]{Philippe Lemey}
\author[6]{Colin Russell}
\author[6,7]{Derek Smith}
\author[1,8]{Andrew Rambaut}

@@

Mutations to the HA1 region of the hemagglutinin (HA) protein are thought to drive the majority of antigenic drift in the influenza virus \cite{Nelson07NatRevGenet}. 
Experimental characterization of antigenic phenotype is possible through the hemagglutination inhibition (HI) assay \cite{Hirst43}, which measures the cross-reactivity of one virus strain to serum raised against another strain. 
Sera from older strains react poorly with more recent viruses resulting in new strains having a transmission advantage over established strains. 
The results of many HI assays across a multitude of virus strains can be combined to yield a two-dimensional map, representing antigenic similarity and distance as an easily visualized and quantified measure \cite{Smith04}. 
The antigenic map of influenza A (H3N2) has shown largely linear movement of the influenza virus population since its introduction in 1968. However, evolution of antigenic phenotype appears punctuated with periods of stasis interspersed by periods of more rapid innovation, while genetic evolution appears more continuous \cite{Smith04}, suggesting that a relatively small number of genetic changes or combinations of genetic changes may drive changes in antigenic phenotype. 
The process of antigenic drift results in the rapid turnover of the virus population. 
Although mutation occurs rapidly, standing genetic diversity is low and phylogenetic analysis shows a characteristically `spindly' tree with a single predominant trunk lineage and transitory side branches that persist for only 1--5 years \cite{Fitch97}.

@@

In order to assess patterns of antigenic evolution among influenza strains, we implemented a Bayesian probabilistic analog of multidimensional scaling, referred to here as BMDS.
In this model, each virus strain is given a $N$ dimensional antigenic location, and the distance between strains in this antigenic space is proportional to their level of cross-reactivity, as measured by the hemagglutination inhibition (HI) assay.
The BMDS model provides an expectation for HI titers, such that a map distance of one antigenic unit between strains translates to a 2-fold drop in HI titer.
A map that produces pairwise distances congruent with the observed titers will have a high likelihood and will be favored by the BMDS model.
We integrate over sources of uncertainty, such as strain location, in flexible Bayesian fashion.

We started with a BMDS implementation of antigenic model used by Smith et al. \cite{Smith04}, where each ferret sera and each virus isolate receive a unique location in a 2D antigenic landscape, and map distance is proportional to drop in HI titer relative to the maximum for a particular ferret sera.
This model performed well, yielding an average absolute predictive error of 0.71 $\mathrm{log}_2$ HI titers on the 6545 training measurements used to build the model, and an average absolute error of 1.23 on 723 test measurements (Table \ref{errortable}).
We extended this model by forcing sera and virus from the same influenza strain to share an antigenic location, reducing the number of required parameters and decreasing test error to 1.05.
We further extended this model by sampling the expected titer for the homologous comparison to the sera strain (column effects) and the expected titer for the homologous comparison to the virus strain (row effects).
We found that adding these column and row effects, decreased training error to 0.67 and test error to 0.87.
We experimented with increasing the number of dimensions beyond two, but found that further increases in dimensionality did not significantly improve test error (Table \ref{errortable}).

\subsection*{Antigenic drift in across influenza types and subtypes}

Through this analysis we find that the antigenic phenotype of influenza A (H3N2) underwent rapid turnover from 1968 to 2011 (Figure~\ref{fluAGrid}A).  
Here, cross-reactive measurements only exist betwen strains sampled at most 11 years apart, leaving only threshold titers, e.g.\ `$<$40', in more temporally distant comparisons.  
Because of the threshold of sensitivity of the HI assay, it's impossible to distinguish a linear trajectory in 2D antigenic space, from a curved trajectory, so long as the curve does not bring antigenic phenotype full circle to have cross-reactive measurements between temporally distant strains.
To solve this problem of identifiability, we enforced a weak prior that favors linear movement in the 2D antigen space.
We find that influenza A (H3N2) moved XXX antigenic units per year (XXX--XXX 95\% credible interval) from 1968 to 2011 (Figure~\ref{fluAGrid}C).
However, the rate of antigenic drift was not constant, with year-to-year movement of the mean antigenic phenotype averaging XXX units, but showing an interquartile range of XXX--XXX.

%%% Figure 1: mds %%%
\begin{figure}[tb]
	\centering		
	\includegraphics[width=\textwidth]{figures/fluAGrid}
	\caption{\textbf{Antigenic locations of influenza H3N2 and H1N1.} (A) and (B) Antigenic maps showing the mean posterior location of 338 strains of H3N2 influenza and 243 strains of H1N1 influenza.  
	The map has been oriented so that the primary axis of variation lies along the $x$-axis (AG1), with the $y$-axis (AG2) orthogonal to this axis.  
	(C) and (D) Antigenic location along the primary axis of variation (AG1) vs.\ year of virus isolation.  
	The dashed lines show the relationship of between time and AG1 with a slope of \tbc{XXX} for H3N2 and \tbc{XXX} for H1N1.  
	(E) and (F) Antigenic location along the secondary axis of variation (AG2) vs.\ year of virus isolation.  
	Antigenic units represent two-fold dilutions of the HI assay, and strains have been colored based on year of isolation.  
	\tbc{Decide whether to just show mean posterior or show distributions.}
	} 
	\label{fluAGrid} 
\end{figure}

\begin{table}[tb]
	\centering
	\caption{\textbf{MDS precision and absolute error of $\mathrm{log}_2$ HI titer prediction for training and test data across models.}}
	\label{errortable}
	\begin{tabular*}{1.0\textwidth}{ c c c c c c c } 
	\hline
	Locations 	& 	Column effects 	&	Row effects	&	MDS		&	Precision	& 	Training error	& 	Test error	\\
	\hline				
	Unique 		&	Fixed 			&	None		&	2D		&	0.79		& 	0.71			&	1.23		\\
	Shared		&	Fixed			&	None		&	2D		&	0.52		&	0.91			&	1.20		\\
	Shared		&	Sampled			&	None		&	2D		&	0.62		&	0.88			&	1.05		\\
	Shared		&	Sampled			&	Sampled		&	2D		&	0.87		&	0.72			&	0.88		\\
	Shared		&	Sampled			&	Sampled		&	3D		&	0.92		&	0.67			&	0.87		\\
	Shared		&	Sampled			&	Sampled		&	4D		&	XXX			&	XXX				&	XXX			\\
	Shared		&	Sampled			&	Sampled		&	5D		&	XXX			&	XXX				&	XXX			\\	
	\hline
	\end{tabular*}
\end{table}

@@

Antigenic data for influenza A (H1N1) was collected from \cite{Faress05,Chakraverty82,Chakraverty86,Cox83,Daniels85,Daum02,Donatelli93,Kendal78,McDonald07,Nakajima79,Nakajima81,Pereira82,Stevens87,Webster79,Raymond86}.

@@

%%% REFERENCES %%%
\bibliographystyle{plos}
\bibliography{flux}
